A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.

### 2020-02-06 paper.tex
keep_tex: true

---
title: A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.
author: Nicholas J Tierney\* (1,2) & Karthik Ram\* (3)
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: no
    includes:
      in_header: preamble.tex
bibliography: references.bib
header-includes: 
  - \linespread{2} 
  - \usepackage[utf8]{inputenc}
---

@@

\* = Both authors contributed equally to the work

@@

Nicholas Tierney (nicholas.tierney@gmail.com) & Karthik Ram (k@karthik.io)

# Abstract {-}

@@

```{r setup, echo = FALSE}

knitr::opts_chunk$set(echo = FALSE,
                      out.width = "100%")

suppressMessages(library(ggrepel))
suppressMessages(library(tibble))
suppressMessages(library(kableExtra))
suppressMessages(library(magick))
suppressMessages(library(ggplot2))
suppressMessages(library(cowplot))
suppressMessages(library(gridExtra))
```

# Introduction {#intro}

> "Data! data! data!" he cried impatiently. "I can't make bricks without clay." - Sherlock Holmes (The Adventure of the Copper Beeches by Sir Arthur Conan Doyle)

Data are a fundamental currency upon which scientific discoveries are made. Without access to good data, it becomes extremely difficult, if not impossible, to advance science. Yet, a large majority of data on which published research papers are based rarely see the light of day and are only visible to the original authors [@Rowhani-Farid2016; @Stodden2018].  Sharing data sets upon publication of a research paper has benefits to individual researchers, often through increased visibility of the work [@Popkin2019; @Kirk2019]. But there is also a component of benefit to the broader scientific community. This primarily comes in the form of potential for reuse in other contexts along use for training and teaching (@McKiernan2016). Assuming that the data have no privacy concerns (e.g., human subjects, locations of critically endangered species), or that the act of sharing does not put the authors at a competitive disadvantage (data can be embargoed for reasonable periods of time), sharing data will always have a net positive benefit. First and foremost, sharing data along with other artifacts can make it easier for a reader to independently reproduce the results, thereby increasing transparency and trust in the work. The act of easy data sharing can also improves model training, as many different models can be tried and tested on latest data sources, closing the loop on research and application of statistical techniques. Existing data sets can be combined or linked with new or existing data, fostering the development and synthesis of new ideas and research areas. The biggest of these benefits is the overall increase in reproducibility.

For nearly two decades, researchers who work in areas related to computational science have pushed for better standards to verify scientific claims, especially in areas where a full replication of the study would be prohibitively expensive. To meet these minimal standards, there must be easy access to the data, models, and code. Among the different areas with a computational bent, the bioinformatics community in particular has a strong culture around open source [@Gentleman2004], and has made releasing code and associated software a recognized mainstream activity. Many journals in these fields have also pushed authors to submit code (in the form of model implementations) alongside their papers, with a few journals going as far as providing a "reproducibility review" [@Peng2011].

**In this paper we focus on the practical side of sharing data for the purpose of reproducibility.** Our goal is to describe various methods in which an analyst can share their data with minimal friction. We steer clear of idealistic ideas such as the FAIR data principles [@Wilkinson2016] since they still do not help a researcher share their data. We also skip the discussion around citation and credit because data citations are still poorly tracked and there is no standardized metric or a h-index equivalent for data as of this writing.

For a piece of computational research to be minimally reproducible, it requires three distinct elements: 1) Code;  2) Computing environment, and 3) Data. The first two of these challenges have largely been solved [@Poisot2019].

Although code sharing in science had a rocky start [@Barnes2010], more and more code writing by scientists is being shared, partly due to the rapid increase in training opportunities made available by organizations like The Carpentries, combined with the rapid adoption of Github by scientists [@Ram2013]. The bigger driver for this may also be connected to the rise in popularity of data science as a discipline distinct from statistics [@Donoho2017]. This rapid growth in data science has largely been fueled by easy access to open source software tools. Programming languages such as Python, R and Julia help scientists implement and share new methods to work with data [@rcore; @python; @julia]. Each of these languages is supported by thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. As of this writing, Python, Julia, and R have  167k packages [@pypi], ~ 14k packages [@cran] and ~2k  packages [@julia-pkgman] respectively. These packages form the building blocks of a data scientists daily workflow.

In a modern computational workflow a user can `pip install` a package in Python, or use `install.packages` in R to install all software dependencies for a project. By relying on the idea of having research compendia [@gentleman2007statistical] or something as simple as a requirements file, it is relatively straightforward to install all the necessary scaffolding. Data on the other hand are rarely accessible that easily.

@@

The computing environment is also easily captured with modern tools such as Docker [@Boettiger2015; @Jupyter2018]. Modern tools such as Binder [@Jupyter2018] can parse Docker files and dependency trees to provide on demand, live notebooks in R and Python that a reader can immediately execute in the browser without dealing with the challenges of local installation. This makes it simple to load a specific environment to run any analysis. Code is handled by version control with tools like Git and GitHub [@git; @github], paired with archiving frameworks such as Zenodo provide access to code (particularly model implementations)[@zenodo]. All the necessary software is available from various package managers (and their numerous geographic mirrors and archives) making it easy to install any version of a software package. However, the biggest challenge, even today, remains easy and repeatable access to data in a data analysis.

Datasets are often far more diverse than code in terms of complexity, size, and formats. This makes them particularly challenging to standardize or easily "install" where the code is running. While there are numerous public and private data repositories, none of them function as package managers, which is what provides so much robustness to code sharing. As a result, data used in an analysis is often read from various locations (local, network, or cloud), various formats, varying levels of tidiness [@Wickham2014]. There is also the overhead associated with data publishing, the act of archiving data in a repository that also mints permanent identifiers, that are not required of all projects due to the effort and time involved. It is worth drawing the distinction between data sharing (making the data available with little effort) and data publishing (archiving the data in a long-term repository, with our without curated metadata).

What aspects of data make them particularly challenging to share from a reproducibility perspective? This is the question we tackle in this paper.  While there are several papers that serve as best-practice guides for formatting data [@Broman2017] and getting them ready for sharing [@Ellis2017], the aims of this paper are a bit different. Our aim is to address the issue of data in the context of reproducibility in data science workflows. In particular we discuss the various tradeoffs one has to consider when documenting and sharing data, when it is worth publishing, and how this would change depending on the use case, perceived impact, and potential audience.

We discuss how to share and/or publish data and cover various tradeoffs when deciding how much to do. We skip detailed discussions of data preparation [@Broman2017; @Ellis2017], or citation (as of this writing, data citation is still in its infancy). We also analyze the state of data contained inside software packages, shared and made available as part of modern data analysis. How often are data shipped inside software packages? How does data format and size impact availability?

# A minimal structure to share data for analysis {#minimal}

To share data analysis data, there should be some minimal set of requirements. For example, the data should contain information on metadata, data dictionaries, the README, and data used in analysis (See Section \@ref(minimal)). No matter where data is submitted, there should ideally be a canonical data repository in one long term archive that links to others. It should also have an accession number, or Digital Object Identification (DOI) number, which allows for it to be cited (discussed further in \@ref(cite)).

@@

1. README: A Human readable description of the data
2. Data dictionary: Human readable dictionary of data contents
3. License: How to use and share the data
4. Citation: How you want your data to be cited
5. Machine readable meta data: Make your data searchable
6. Raw data: The original/first data provided
7. Scripts: To clean raw data ready for analysis
8. Analysis ready data: Final data used in analysis

One basic suggested directory layout is given below in Figure \@ref(fig:code-directory).

```{r code-directory, out.width = "75%", fig.align = "center", fig.cap = "Example directory layout and structure for a data repository."}
knitr::include_graphics(here::here("figures",
                                   "directory-structure.png"))
```

@@

1. README
2. Data dictionary
3. License
4. Citation
5. Analysis ready data
6. Scripts to tidy analysis raw data into ready data

@@

## README: A Human readable description of the data {#readme}

In the context of datasets, READMEs are particularly useful when there are no reliable standards. A README is often the first place people will go to learn more about anything in a folder - they are very common in software, and historically were included so the uppercase letters of README meant it would be at the top of a directory. The README is meant for someone to read and understand more about the data and contains the "who, what, when, where, why, & how":

- **Who** collected it
- **What** is the data
- **When** was it collected
- **Where** was it collected
- **Why** it was collected
- **How** is was collected

@@

Saving a README with the extension `.md` file gives the author the formatting benefits of `markdown`, making it simple to insert links, tables, and make lists. In systems like GitHub, a README file is detected and rendered in nice HTML on the repository by default.

## Data dictionary: Human readable dictionary of data contents {#data-dictionary}

@@

- **variable names**
- **variable labels**
- **variable codes**, and
- special values for **missing data**.

Figure \@ref(fig:variables) shows an example of the data and data dictionary details.
**Variable names** should be short, descriptive names with no spaces or special characters. In the most common tabular data case, these correspond to column names, for example, "job_position", "faculty_level", and "years_at_uni". **Variable labels** are longer descriptions of variables. For example "University Job Position", "University Faculty Position Level", and  "Number of Years Worked at University" [@McGill-codebook; @Arslan2019]. **Variable codes** apply to categorical (factor) variables, and are the values for their contents. For example, a variable could contain answers to a question, with values and codes like so: `0 = no, 1 = yes`, to indicate if someone is in statistics, for example.  These should be consistent across similar variables to avoid problems where `0 = yes` for one variable, and `1 = yes` in another. Date variables should have consistent formatting. For example, all date information could be in format "YYYY-MM-DD", and this should not be mixed with "YYYY-DD-MM". Date formatting should be described in the variable labels. **Missing data** are values that should have been observed, but were not. The code for missingness should be documented in the data dictionary, and should nominally be `NA`. If the reason for missingness is known it should be recorded. For example, censored data, patient drop out, or measurement error can have different values, such as "unknown", -99, or other value codes [@White2013; @Broman2017].

```{r variables, fig.cap = "Four tables - the data, and it's variable names, variable codes, and the meaning of missing values", fig.height = 8}

dat <- tribble(
  ~job_position, ~outside, ~statistics, ~date_started,
   "Lecturer",        1,        1,      "2019-01-01",
   "Professor",       0,        0,      NA,
   "Senior Lecturer", 1,        1,      "unknown"
)

variable_labels <- tribble(
  ~variable, ~description,
  "job_position", "University Job Position",
  "outside",      "Did this person study outside this university?",
  "statistics",   "Is this person in the statistics faculty",
  "date_started", "What date did this perosn start their position?"
)

variable_codes <- tribble(
  ~variable,        ~code,
   "job_position",   "text description of job position",
   "outside",       "0 = not studied outside, \n1 = has studied outside",
   "statistics",    "0 = is not in statistics dept., \n1 = is currently in statistics dept.",
   "date_started",   "date variable in format YYYY-MM-DD, \ne.g., 2001-01-01 is 1st January, 2001"
)

missing_values <- tribble(
  ~value, ~meaning,
  NA,     "Missing value",
  "unknown", "Missing value"
)

grid.arrange(
  tableGrob(dat),
  tableGrob(variable_labels),
  tableGrob(variable_codes),
  tableGrob(missing_values),
  ncol = 1
)

```

Table \@ref(tab:table-data-incarcerate) shows an example data dictionary table from the Tidy Tuesday repository on incarceration trends [@tidy-tuesday-incarcerate]. This includes information on the variable, its class (type), and a longer description.

```{r table-data-incarcerate}
dictionary_incarcerate <- data.frame(
  Variable = c("year", "urbanicity", "pop_category", "rate_per_100000"),
  Class  = c("integer (date)",
             "character",
             "character",
             "double"),
  Description = c("Year",
                  "County-type (urban, suburban, small/mid, rural)",
                  "Category for population - either race, gender, or Total",
                  "Rate within a category for prison population per 100,000 people")
)

knitr::kable(dictionary_incarcerate,
             caption = "The prisoner summary data dictionary, with columns for the variable, its class, and a short description of the contents of the variable.",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))
```

Data dictionaries should be placed in the README and presented as a table. Every data dictionary should also be provided in its raw form (e.g., a CSV) in the repository, so they aren't "trapped" in the README.

## License: How to use and share the data {#license}

@@

1. Creative Commons Attribution 4.0 International Public License (CC BY), and
2. Creative Commons CC0 1.0 Universal (CC0)

Other licenses or notices to be aware of are **copyrighted data**, and **data embargoes**. If you are working with data **already copyrighted**, (for example under CC BY or CC0), you must give follow appropriate guidelines for giving credit. Data may also be under **embargo**. This means data cannot be shared more widely until a specific release time. If sharing data under an embargo, include detailed information on the embargo requirements in: the README, and in separate correspondence with those who receive the data. Databases may also be licensed with the Open Data Commons Open Database License: https://opendatacommons.org/licenses/odbl/, which provides provisions for sharing, creating, and adapting, provided that work is attributed, shared, and kept open.

@@

### CC BY

@@

* You must provide appropriate credit to the source. This means listing the names of the creators.
* Link back to the CC BY license, and
* Clearly show if changes were made.
* Data cannot be sub-licensed, that is - a change to the existing license
* There is also no warranty, so the person or people who obtained the data cannot be held liable.

The journal PLOS Comp Bio requires that data submitted cannot be more restrictive than CC BY [@plos-comp-bio-data]. For a brief overview of the CC BY, suitable to include in a README, see @ccby-short. For the full license, see @ccby-long.

### CC0

The CC0 is a "public domain" license. Data with a CC0 license means the data owners waive all their rights to the work, and it now "owned" by the public. The data can be freely shared, copied, modified, and distributed, even for commercial purposes _without asking permission_. When using data with CC0, it is good practice to cite the original paper, but it is not required. If you wish to use the CC0, see @choose-cc0.  For a brief overview of the CC0, see @cc0-short, and for the full license, see @cc0-long.

## Citation: How you want your data to be cited {#cite}

A Digital Object Identifier (DOI), is a unique identifier for a digital object such as a paper, poster, or software, and is a prerequisite for citation. For example, if the data are deposited in repositories like Dryad, Zenodo, or the Open Science Framework, the best practice would be to copy the citation created by these repositories. Under the hood, the DOI is "minted" by DataCite, who provide DOIs for research outputs and data. This means that when citing data, it only makes sense to cite datasets that have been deposited into a DataCite compliant repository. If a DOI is unavailable, a citation will be meaningless, as it cannot be tracked by any means. A file named `citation` should be placed in the directory, at the same level as the `README`. It should contain a DOI, and could be in `.bibtex` format.

## Machine readable metadata {#machine-metadata}

The README and data dictionary provides _human readable_ information on the data. To ensure data types are preserved - for example, dates are dates, names are characters - there needs to be some form of _machine readable_ metadata. This provides a structure allowing the data to be indexed and searched online, through services such as google datasets search [@google-data-search]. An excellent standard for metadata is Table Schema written by frictionless data [@Fowler2017]. For example, a dataset "demographics" with three variables is shown in Table \@ref(tab:table-machine-readable), and the Java Script Object Notation (JSON) equivalent in Figure \@ref(fig:json-code).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.},
  pdfauthor={Nicholas J Tierney* (1,2) \& Karthik Ram* (3)},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
%% Any special functions or other packages can be loaded here.

%% FIGURE FLOATING taken from
%% https://stackoverflow.com/questions/16626462/figure-position-in-markdown-when-converting-to-pdf-with-knitr-and-pandoc
\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}

%% MONASH STUFF TAKEN FROM ROB

%% SPACING
\RequirePackage{setspace}
\spacing{1.5}

%% HEADERS AND FOOTERS
\RequirePackage{fancyhdr}
\pagestyle{fancy}
\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
\makeatletter
\lhead{\textsf{\expandafter{\@title}}}
\makeatother
\rhead{}
\cfoot{}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}
\fancypagestyle{plain}{%
\fancyhf{} % clear all header and footer fields
\fancyfoot[C]{\sffamily\thepage} % except the center
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

%% LINE AND PAGE BREAKING
\sloppy
\clubpenalty = 10000
\widowpenalty = 10000
\brokenpenalty = 10000
\RequirePackage{microtype}

%% PARAGRAPH BREAKS
\setlength{\parskip}{1.4ex}
\setlength{\parindent}{0em}

%% CAPTIONS
\RequirePackage{caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}
\linespread{2}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.}
\author{Nicholas J Tierney* (1,2) \& Karthik Ram* (3)}
\date{}

\begin{document}
\maketitle

@@

* = Both authors contributed equally to the work

@@

Nicholas Tierney (\href{mailto:nicholas.tierney@gmail.com}{\nolinkurl{nicholas.tierney@gmail.com}}) \& Karthik Ram (\href{mailto:k@karthik.io}{\nolinkurl{k@karthik.io}})

\hypertarget{abstract}{%
\section*{Abstract}\label{abstract}}
\addcontentsline{toc}{section}{Abstract}

@@

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

\begin{quote}
``Data! data! data!'' he cried impatiently. ``I can't make bricks without clay.'' - Sherlock Holmes (The Adventure of the Copper Beeches by Sir Arthur Conan Doyle)
\end{quote}

Data are a fundamental currency upon which scientific discoveries are made. Without access to good data, it becomes extremely difficult, if not impossible, to advance science. Yet, a large majority of data on which published research papers are based rarely see the light of day and are only visible to the original authors (Rowhani-Farid and Barnett 2016; Stodden, Seiler, and Ma 2018). Sharing data sets upon publication of a research paper has benefits to individual researchers, often through increased visibility of the work (Popkin 2019; Kirk and Norton 2019). But there is also a component of benefit to the broader scientific community. This primarily comes in the form of potential for reuse in other contexts along use for training and teaching (McKiernan et al. (2016)). Assuming that the data have no privacy concerns (e.g., human subjects, locations of critically endangered species), or that the act of sharing does not put the authors at a competitive disadvantage (data can be embargoed for reasonable periods of time), sharing data will always have a net positive benefit. First and foremost, sharing data along with other artifacts can make it easier for a reader to independently reproduce the results, thereby increasing transparency and trust in the work. The act of easy data sharing can also improves model training, as many different models can be tried and tested on latest data sources, closing the loop on research and application of statistical techniques. Existing data sets can be combined or linked with new or existing data, fostering the development and synthesis of new ideas and research areas. The biggest of these benefits is the overall increase in reproducibility.

For nearly two decades, researchers who work in areas related to computational science have pushed for better standards to verify scientific claims, especially in areas where a full replication of the study would be prohibitively expensive. To meet these minimal standards, there must be easy access to the data, models, and code. Among the different areas with a computational bent, the bioinformatics community in particular has a strong culture around open source (Gentleman et al. 2004), and has made releasing code and associated software a recognized mainstream activity. Many journals in these fields have also pushed authors to submit code (in the form of model implementations) alongside their papers, with a few journals going as far as providing a ``reproducibility review'' (Peng 2011).

\textbf{In this paper we focus on the practical side of sharing data for the purpose of reproducibility.} Our goal is to describe various methods in which an analyst can share their data with minimal friction. We steer clear of idealistic ideas such as the FAIR data principles (Wilkinson et al. 2016) since they still do not help a researcher share their data. We also skip the discussion around citation and credit because data citations are still poorly tracked and there is no standardized metric or a h-index equivalent for data as of this writing.

For a piece of computational research to be minimally reproducible, it requires three distinct elements: 1) Code; 2) Computing environment, and 3) Data. The first two of these challenges have largely been solved (Poisot et al. 2019).

Although code sharing in science had a rocky start (Barnes 2010), more and more code writing by scientists is being shared, partly due to the rapid increase in training opportunities made available by organizations like The Carpentries, combined with the rapid adoption of Github by scientists (Ram 2013). The bigger driver for this may also be connected to the rise in popularity of data science as a discipline distinct from statistics (Donoho 2017). This rapid growth in data science has largely been fueled by easy access to open source software tools. Programming languages such as Python, R and Julia help scientists implement and share new methods to work with data (R Core Team 2019; ``Python,'' n.d.; Bezanson et al. 2017). Each of these languages is supported by thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. As of this writing, Python, Julia, and R have 167k packages (``Pypi,'' n.d.), \textasciitilde{} 14k packages (``Cran,'' n.d.) and \textasciitilde2k packages (``Julia-Pkgman,'' n.d.) respectively. These packages form the building blocks of a data scientists daily workflow.

In a modern computational workflow a user can \texttt{pip\ install} a package in Python, or use \texttt{install.packages} in R to install all software dependencies for a project. By relying on the idea of having research compendia (Gentleman and Temple Lang 2007) or something as simple as a requirements file, it is relatively straightforward to install all the necessary scaffolding. Data on the other hand are rarely accessible that easily.

@@

The computing environment is also easily captured with modern tools such as Docker (Boettiger 2015; Jupyter 2018). Modern tools such as Binder (Jupyter 2018) can parse Docker files and dependency trees to provide on demand, live notebooks in R and Python that a reader can immediately execute in the browser without dealing with the challenges of local installation. This makes it simple to load a specific environment to run any analysis. Code is handled by version control with tools like Git and GitHub (``Git,'' n.d.; ``Github,'' n.d.), paired with archiving frameworks such as Zenodo provide access to code (particularly model implementations)(Zenodo 2016). All the necessary software is available from various package managers (and their numerous geographic mirrors and archives) making it easy to install any version of a software package. However, the biggest challenge, even today, remains easy and repeatable access to data in a data analysis.

Datasets are often far more diverse than code in terms of complexity, size, and formats. This makes them particularly challenging to standardize or easily ``install'' where the code is running. While there are numerous public and private data repositories, none of them function as package managers, which is what provides so much robustness to code sharing. As a result, data used in an analysis is often read from various locations (local, network, or cloud), various formats, varying levels of tidiness (Wickham 2014). There is also the overhead associated with data publishing, the act of archiving data in a repository that also mints permanent identifiers, that are not required of all projects due to the effort and time involved. It is worth drawing the distinction between data sharing (making the data available with little effort) and data publishing (archiving the data in a long-term repository, with our without curated metadata).

What aspects of data make them particularly challenging to share from a reproducibility perspective? This is the question we tackle in this paper. While there are several papers that serve as best-practice guides for formatting data (Broman and Woo 2017) and getting them ready for sharing (Ellis and Leek 2017), the aims of this paper are a bit different. Our aim is to address the issue of data in the context of reproducibility in data science workflows. In particular we discuss the various tradeoffs one has to consider when documenting and sharing data, when it is worth publishing, and how this would change depending on the use case, perceived impact, and potential audience.

We discuss how to share and/or publish data and cover various tradeoffs when deciding how much to do. We skip detailed discussions of data preparation (Broman and Woo 2017; Ellis and Leek 2017), or citation (as of this writing, data citation is still in its infancy). We also analyze the state of data contained inside software packages, shared and made available as part of modern data analysis. How often are data shipped inside software packages? How does data format and size impact availability?

\hypertarget{minimal}{%
\section{A minimal structure to share data for analysis}\label{minimal}}

To share data analysis data, there should be some minimal set of requirements. For example, the data should contain information on metadata, data dictionaries, the README, and data used in analysis (See Section \ref{minimal}). No matter where data is submitted, there should ideally be a canonical data repository in one long term archive that links to others. It should also have an accession number, or Digital Object Identification (DOI) number, which allows for it to be cited (discussed further in \ref{cite}).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  README: A Human readable description of the data
\item
  Data dictionary: Human readable dictionary of data contents
\item
  License: How to use and share the data
\item
  Citation: How you want your data to be cited
\item
  Machine readable meta data: Make your data searchable
\item
  Raw data: The original/first data provided
\item
  Scripts: To clean raw data ready for analysis
\item
  Analysis ready data: Final data used in analysis
\end{enumerate}

One basic suggested directory layout is given below in Figure \ref{fig:code-directory}.

\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{/Users/ntie0001/github/njtierney/ddd/figures/directory-structure}

}

\caption{Example directory layout and structure for a data repository.}\label{fig:code-directory}
\end{figure}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  README
\item
  Data dictionary
\item
  License
\item
  Citation
\item
  Analysis ready data
\item
  Scripts to tidy analysis raw data into ready data
\end{enumerate}

@@

\hypertarget{readme}{%
\subsection{README: A Human readable description of the data}\label{readme}}

In the context of datasets, READMEs are particularly useful when there are no reliable standards. A README is often the first place people will go to learn more about anything in a folder - they are very common in software, and historically were included so the uppercase letters of README meant it would be at the top of a directory. The README is meant for someone to read and understand more about the data and contains the ``who, what, when, where, why, \& how'':

\begin{itemize}
\tightlist
\item
  \textbf{Who} collected it
\item
  \textbf{What} is the data
\item
  \textbf{When} was it collected
\item
  \textbf{Where} was it collected
\item
  \textbf{Why} it was collected
\item
  \textbf{How} is was collected
\end{itemize}

@@

Saving a README with the extension \texttt{.md} file gives the author the formatting benefits of \texttt{markdown}, making it simple to insert links, tables, and make lists. In systems like GitHub, a README file is detected and rendered in nice HTML on the repository by default.

\hypertarget{data-dictionary}{%
\subsection{Data dictionary: Human readable dictionary of data contents}\label{data-dictionary}}

@@

\begin{itemize}
\tightlist
\item
  \textbf{variable names}
\item
  \textbf{variable labels}
\item
  \textbf{variable codes}, and
\item
  special values for \textbf{missing data}.
\end{itemize}

Figure \ref{fig:variables} shows an example of the data and data dictionary details.
\textbf{Variable names} should be short, descriptive names with no spaces or special characters. In the most common tabular data case, these correspond to column names, for example, ``job\_position'', ``faculty\_level'', and ``years\_at\_uni''. \textbf{Variable labels} are longer descriptions of variables. For example ``University Job Position'', ``University Faculty Position Level'', and ``Number of Years Worked at University'' (``Codebook Cookbook: A Guide to Writing a Good Codebook for Data Analysis Projects in Medicine'' n.d.; Arslan 2019). \textbf{Variable codes} apply to categorical (factor) variables, and are the values for their contents. For example, a variable could contain answers to a question, with values and codes like so: \texttt{0\ =\ no,\ 1\ =\ yes}, to indicate if someone is in statistics, for example. These should be consistent across similar variables to avoid problems where \texttt{0\ =\ yes} for one variable, and \texttt{1\ =\ yes} in another. Date variables should have consistent formatting. For example, all date information could be in format ``YYYY-MM-DD'', and this should not be mixed with ``YYYY-DD-MM''. Date formatting should be described in the variable labels. \textbf{Missing data} are values that should have been observed, but were not. The code for missingness should be documented in the data dictionary, and should nominally be \texttt{NA}. If the reason for missingness is known it should be recorded. For example, censored data, patient drop out, or measurement error can have different values, such as ``unknown'', -99, or other value codes (White et al. 2013; Broman and Woo 2017).

\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/variables-1} \caption{Four tables - the data, and it's variable names, variable codes, and the meaning of missing values}\label{fig:variables}
\end{figure}

Table \ref{tab:table-data-incarcerate} shows an example data dictionary table from the Tidy Tuesday repository on incarceration trends (``Tidy-Tuesday-Incarcerate,'' n.d.). This includes information on the variable, its class (type), and a longer description.

\begin{table}[!h]

\caption{\label{tab:table-data-incarcerate}The prisoner summary data dictionary, with columns for the variable, its class, and a short description of the contents of the variable.}
\centering
\begin{tabular}[t]{lll}
\toprule
Variable & Class & Description\\
\midrule
year & integer (date) & Year\\
urbanicity & character & County-type (urban, suburban, small/mid, rural)\\
pop\_category & character & Category for population - either race, gender, or Total\\
rate\_per\_100000 & double & Rate within a category for prison population per 100,000 people\\
\bottomrule
\end{tabular}
\end{table}

Data dictionaries should be placed in the README and presented as a table. Every data dictionary should also be provided in its raw form (e.g., a CSV) in the repository, so they aren't ``trapped'' in the README.

\hypertarget{license}{%
\subsection{License: How to use and share the data}\label{license}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Creative Commons Attribution 4.0 International Public License (CC BY), and
\item
  Creative Commons CC0 1.0 Universal (CC0)
\end{enumerate}

Other licenses or notices to be aware of are \textbf{copyrighted data}, and \textbf{data embargoes}. If you are working with data \textbf{already copyrighted}, (for example under CC BY or CC0), you must give follow appropriate guidelines for giving credit. Data may also be under \textbf{embargo}. This means data cannot be shared more widely until a specific release time. If sharing data under an embargo, include detailed information on the embargo requirements in: the README, and in separate correspondence with those who receive the data. Databases may also be licensed with the Open Data Commons Open Database License: \url{https://opendatacommons.org/licenses/odbl/}, which provides provisions for sharing, creating, and adapting, provided that work is attributed, shared, and kept open.

@@

\hypertarget{cc-by}{%
\subsubsection{CC BY}\label{cc-by}}

@@

\begin{itemize}
\tightlist
\item
  You must provide appropriate credit to the source. This means listing the names of the creators.
\item
  Link back to the CC BY license, and
\item
  Clearly show if changes were made.
\item
  Data cannot be sub-licensed, that is - a change to the existing license
\item
  There is also no warranty, so the person or people who obtained the data cannot be held liable.
\end{itemize}

The journal PLOS Comp Bio requires that data submitted cannot be more restrictive than CC BY (``PLOS Computational Biology,'' n.d.). For a brief overview of the CC BY, suitable to include in a README, see (``CCBY Short Guide,'' n.d.). For the full license, see (``CCBY Full License,'' n.d.).

\hypertarget{cc0}{%
\subsubsection{CC0}\label{cc0}}

The CC0 is a ``public domain'' license. Data with a CC0 license means the data owners waive all their rights to the work, and it now ``owned'' by the public. The data can be freely shared, copied, modified, and distributed, even for commercial purposes \emph{without asking permission}. When using data with CC0, it is good practice to cite the original paper, but it is not required. If you wish to use the CC0, see (``Choose-Cc0,'' n.d.). For a brief overview of the CC0, see (``CC0 Short Guide,'' n.d.), and for the full license, see (``CC0 Full License,'' n.d.).

\hypertarget{cite}{%
\subsection{Citation: How you want your data to be cited}\label{cite}}

A Digital Object Identifier (DOI), is a unique identifier for a digital object such as a paper, poster, or software, and is a prerequisite for citation. For example, if the data are deposited in repositories like Dryad, Zenodo, or the Open Science Framework, the best practice would be to copy the citation created by these repositories. Under the hood, the DOI is ``minted'' by DataCite, who provide DOIs for research outputs and data. This means that when citing data, it only makes sense to cite datasets that have been deposited into a DataCite compliant repository. If a DOI is unavailable, a citation will be meaningless, as it cannot be tracked by any means. A file named \texttt{citation} should be placed in the directory, at the same level as the \texttt{README}. It should contain a DOI, and could be in \texttt{.bibtex} format.

\hypertarget{machine-metadata}{%
\subsection{Machine readable metadata}\label{machine-metadata}}
